A thermodynamic view of dusty protoplanetary disks

### 2017-08-08 paper.tex
final revision before submission, spell-checked

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follow the gas flow.
  %AY: Edited b/c "This is" was too vague
  This near conservation of dust-to-gas ratio along streamlines
  is analogous to the near conservation of entropy along flows of (dust-free) gas with weak heating and cooling.   
  We develop this thermodynamic analogy into a
  framework to study dusty gas dynamics in protoplanetary disks. We  
  show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this correspondence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilties; 
\item 
 radial dust edges are unstable if the dust is vertically well-mixed;  
\item 
  the streaming instability neccessarily involves a gas pressure
  response 
  that lags behind dust density; 
%  the streaming instability neccessarily involves a phase lag between
%  gas pressure and dust densities; 
\item dust-loading introduces buoyancy forces that generally 
  stabilizes the vertical shear 
  instability associated with global radial temperature gradients. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss %future applications of this thermodynamic framework in
%finding 
 dusty analogs of other hydrodynamic 
processes (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and how to simulate dusty protoplanetary disks
with minor tweaks to existing codes for pure gas dynamics.%, as well as possible generalizations
%of our analogy between dust-laden and nearly adiabatic flows.

@@

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics coupled to explicit
Lagragian dust particles
\citep{johansen2006, nelson10,bai10,yang14,zhu14,gibbons15,simon16,baruteau16}. 
The equation of motion for each particle is solved 
including dust-gas drag; the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

\begin{figure}
\tikzstyle{process} = [rectangle, minimum width=2cm, minimum height=1cm, text centered, draw=black, fill=orange!30]
  \tikzstyle{line} = [draw, -triangle 45,thick]  
  \begin{center}
    \begin{tikzpicture}[node distance=5cm]
      \node[process](ellip){isothermal gas+dust+drag};
      \node[process,right of=ellip, node distance = 5.5cm](gi){pure 
        gas+cooling};
      \path [line](ellip) -- node [yshift=.8em]
      {\textcolor{blue}{equivalent} for} (gi);
      \path [line](gi) -- node [yshift=-.8em] {small particles} (ellip);
    \end{tikzpicture}
 \end{center}
\caption{The central concept of this work. A mixture of small particles
  coupled to isothermal gas behave similarly to an adiabatic, pure gas
  subject to cooling.  This
  permits one to use the equations for standard hydrodynamics to study
  dust-gas dynamics. \label{concept_chart}}
\end{figure}

@@

\section{Growing oscillations by doing work}\label{grow_osc}
A general result in fluid dynamics is that work is done whenever
oscillations in the pressure and density of the fluid are not in
phase. If  the average work done is positive, then oscillation
amplitudes grow: the work done allows the system to `overshoot' beyond 
the amplitude of preceeding oscillations.

@@

From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure is still 
increasing. This over-compensation will cause the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
is still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing when 
descending.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/drag}\\
\includegraphics[width=\linewidth]{figures/pdv}
  \caption{Thermodynamic interpretation of growing oscillations in a
    fluid. Top: pressure and density evolution in time. 
    Bottom: oscillation cycle in the P-V plane. In the case shown, 
    dust-drag causes 
    pressure to lag behind density, which results in a clockwise path in
    the P-V plane, implying positive work done by the fluid, which 
     would increase oscillation amplitudes.  
    \label{pdv_cartoon}
  }
\end{figure}

@@

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not neccessarily in phase, because dust particles can
drift relative to the gas.   
If this causes $P$ to lag 
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

@@

\input{model}
\input{linear}
\input{result}
\input{summary}

%{\bf acknowledgements. s. inutsuka for asking about kz=0 limit}

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follow the gas flow.
  %AY: Edited b/c "This is" was too vague
  This near conservation of dust-to-gas ratio along streamlines
  is analogous to the near conservation of entropy along flows of (dust-free) gas with weak heating and cooling.   
  We develop this thermodynamic analogy into a
  framework to study dusty gas dynamics in protoplanetary disks. We  
  show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this correspondence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilities; 
\item 
 radial dust edges are unstable if the dust is vertically well-mixed;  
\item 
  the streaming instability necessarily involves a gas pressure
  response 
  that lags behind dust density; 
%  the streaming instability neccessarily involves a phase lag between
%  gas pressure and dust densities; 
\item dust-loading introduces buoyancy forces that generally 
  stabilizes the vertical shear 
  instability associated with global radial temperature gradients. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss %future applications of this thermodynamic framework in
%finding 
 dusty analogs of other hydrodynamic 
processes (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and how to simulate dusty protoplanetary disks
with minor tweaks to existing codes for pure gas dynamics.%, as well as possible generalizations
%of our analogy between dust-laden and nearly adiabatic flows.

@@

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics coupled to explicit
Lagrangian dust particles
\citep{johansen2006, nelson10,bai10,yang14,zhu14,gibbons15,simon16,baruteau16}. 
The equation of motion for each particle is solved 
including dust-gas drag; the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

\begin{figure}
\tikzstyle{process} = [rectangle, minimum width=2cm, minimum height=1cm, text centered, draw=black, fill=orange!30]
  \tikzstyle{line} = [draw, -triangle 45,thick]  
  \begin{center}
    \begin{tikzpicture}[node distance=5cm]
      \node[process](ellip){isothermal gas+dust+drag};
      \node[process,right of=ellip, node distance = 5.5cm](gi){pure 
        gas+cooling};
      \path [line](ellip) -- node [yshift=.8em]
      {\textcolor{blue}{equivalent} for} (gi);
      \path [line](gi) -- node [yshift=-.8em] {small particles} (ellip);
    \end{tikzpicture}
 \end{center}
\caption{The central concept of this work. A mixture of small particles
  imperfectly coupled to isothermal gas behave similarly to an adiabatic, pure gas
  subject to cooling.  This
  permits one to use the equations for standard hydrodynamics to study
  dust-gas dynamics. \label{concept_chart}}
\end{figure}

@@

\section{Growing oscillations by doing work}\label{grow_osc}
A general result in fluid dynamics is that work is done whenever
oscillations in the pressure and density of the fluid are not in
phase. If  the average work done is positive, then oscillation
amplitudes grow: the work done allows the system to `overshoot' beyond 
the amplitude of preceding oscillations.

@@

From $A$ to $B$, a fluid parcel is expanding to return to 
equilibrium, but pressure is still 
increasing. This over-compensation will cause the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
is still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing when 
descending.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/drag}\\
\includegraphics[width=\linewidth]{figures/pdv}
  \caption{Thermodynamic interpretation of growing oscillations in a
    fluid. Top: pressure and density evolution in time. 
    Bottom: oscillation cycle in the P-V plane. In the case shown, 
    dust-drag causes  
    pressure (due to gas only) to lag behind total density. This results in a clockwise path in
    the P-V plane, implying positive work done by the fluid, which 
     would increase oscillation amplitudes.  
    \label{pdv_cartoon}
  }
\end{figure}

@@

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not necessarily in phase, because dust particles can
drift relative to the gas.   
If this causes $P$ to lag 
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

@@

\input{model}
\input{linear}
\input{result}
\input{summary}
\acknowledgements
We thank P. Loren-Aguilar for initial discussions that motivated this study. We also thank 
C. Baruteau, R. Dong, J. Fung, S. Inutsuka, W. Lyra, and M. Pessah 
for comments during the course of this project. This work is supported by the Theoretical Institute for Advanced Research in Astrophysics 
(TIARA) based in Academia Sinica Institute of Astronomy and Astrophysics (ASIAA) and NASA Astrophysics Theory Program grant NNX17AK59G.

### 2017-08-08 result.tex
final revision before submission, spell-checked

\section{
  Dusty vertical shear instability}\label{results}

We now present numerical solutions for vertically stratified,
non-self-gravitating disks, assuming perfectly coupled dust. 
We formally take $\tstop=0$ so the equilibria defined in \S\ref{eqm} 
are exact steady states. In reality,   
%\subsection{Validity}
%{\bf maybe incorporate into intro to this section?}
%We considered $\tstop=0$ in order to perform stability analysis on an
%exact steady state. 
%When $\tstop\neq0$, a stratified dusty 
%disk initialized as in \S\ref{eqm} would evolve 
%as 
dust settles to the midplane on a timescale $t_\mathrm{settle}\sim 
1/\tstop\OmK^2$ \citep{takeuchi02}. However, we expect the 
perfectly-coupled limit to be valid  
provided timescales of interest $t_\mathrm{grow}\ll
t_\mathrm{settle}$. For the VSI this translates to 
%Since VSI growth rates are $O(h_\mathrm{g}\OmK)$, 
%we require  
%\begin{align}
 $ \tstop\OmK \ll h_\mathrm{g}. $
%\end{align} 
For thin PPDs, this is satisfied for $\tstop\OmK \ll O(10^{-2})$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixHd} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disk with midplane dust-to-gas ratio
    of $\epsilon_0=10^{-3}$ (black) and $\epsilon_0=1$ (blue). 
    The dust layer thickness is fixed to $\Hd=0.99\Hg$ so the 
    dust-to-gas ratio $\epsilon$ is approximately constant with
    height. 
    \label{compare_vshear_fixHd}
    }
\end{figure}

@@

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably, surface modes, which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$ (i.e. $\epsilon_0> \hgas$).   
The body modes' growth rates remain $ O(h_\mathrm{g}\OmK)$ 
but their oscillation frequency increases with
dust-loading, i.e. it increases with the vertical buoyancy. 
The total number of modes do not change 
significantly. This is in contrast with the effect of increasing
cooling times in an adiabatic gas disk, for which \citetalias{lin15}
find fewer unstable modes. 
%until the system is eventually completely stable.

@@

We analyze two disks with $\Hd=0.1\Hg$ and 
$\Hd=0.99\Hg$. Fig. \ref{compare_vshear_fixZ} compares the vertical
vhear rate and buoyancy frequency. For $|z|\gtrsim 0.4\Hg$ the two
disks have the same profile with vertical shear dominating over
buoyancy. We thus expect perturbations away from the disk midplane in 
both cases. For $|z|\lesssim 0.4\Hg$, a thin dust 
layer with $\Hd=0.1\Hg$ boosts the vertical shear rate, but the
associated buoyancy is larger still, implying the mid-plane should be
stable. 
%Thus we expect the mid-plane of 
%the disk with $\Hd=0.1\Hg$ to have limited vertical motions.

@@

Fig. \ref{result2d_fixZ} compares the fastest growing VSI body modes  
with $k_x\Hg=30$ for the two cases above.\citepalias[The thinner domain
  adopted here eliminates surface modes, ][]{lin15}.   
We find very similar mode 
frequencies 
\begin{align*}
  \sigma = \begin{cases}
    \left(0.3053\ii - 0.8142\right)h_\mathrm{g}\OmK & \Hd=0.99\Hg, \\
    \left(0.3178\ii - 1.2237\right)h_\mathrm{g}\OmK & \Hd=0.1\Hg,
  \end{cases}
\end{align*}
since the vertical shear profile is similar throughout most of the
disk. However, meridional motions are surpressed near the midplane of
the $\Hd=0.1\Hg$ disk, as expected from the larger buoyancy frequency
relative to vertical shear there. This leads to a structure
analogous to PPD dead zones: a quiescent midplane between active
surface layers \citep{gammie96}.

@@

To connect with previous studies, here  we also use 
the Richardson number $\rich \equiv N_z^2/\left(r\p_z\Omega\right)^2$
to label calculations \citep{youdin02}.   
Numerical simulations show    
non-axisymmetric instabilities develops when $\rich\lesssim0.1$, brought about
by very thin dust layers \citep{chiang08, lee10}. We show that \emph{axisymmetric}
instabilities never develop, however small $\rich$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_thindust} 
  \caption{Maximum VSI growth rate for ultra-thin dust layers $\Hd
    \leq 0.01\Hg$. Vertical shear is dominated by that due to
    dust-settling, but axisymmetric instability is associated with the
    radial temperature gradient $q$. Here, $\rich_0$ is the minium
    Richardson number in the domain in the limit $q\to0$. 
    \label{ultra_thin}
    }
\end{figure}

@@

In \S\ref{iso_perfect} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.  
To demonstate this numerically we consider the
same disk as above but with $q=0$ and choose $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. Then vertical shear $\p_z\Omega$ is 
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).

We show unstable modes for this disk in
Fig. \ref{vert_mixed_modes}. In order to obtain appreciable growth   
rates, we needed to consider super-Solar metalicity $Z=0.03$ and high
radial wavenumber $k_x\Hg=1800$. As expected from the discussion in
\S\ref{iso_perfect}, the disk admits purely growing modes with
$\omega=0$. This is distinct from the growing oscillations associated
with classic VSI discussed above.

>>>>

\section{
  Vertical shear instability with dust}\label{results}

We now present numerical solutions for vertically stratified,
non-self-gravitating disks, assuming perfectly coupled dust. 
We formally take $\tstop=0$ so the equilibrium defined in \S\ref{eqm} 
are exact steady states. In reality,   
%\subsection{Validity}
%{\bf maybe incorporate into intro to this section?}
%We considered $\tstop=0$ in order to perform stability analysis on an
%exact steady state. 
%When $\tstop\neq0$, a stratified dusty 
%disk initialized as in \S\ref{eqm} would evolve 
%as 
dust settles to the midplane on a timescale $t_\mathrm{settle}\sim 
1/\tstop\OmK^2$ \citep{takeuchi02}. However, we expect the 
perfectly-coupled limit to be valid  
provided timescales of interest $t_\mathrm{grow}\ll
t_\mathrm{settle}$. For the VSI this translates to 
%Since VSI growth rates are $O(h_\mathrm{g}\OmK)$, 
%we require  
%\begin{align}
 $ \tstop\OmK \ll h_\mathrm{g}. $
%\end{align} 
For thin PPDs, this is satisfied for $\tstop\OmK \ll O(10^{-2})$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixHd} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disk with midplane dust-to-gas ratio
    of $\epsilon_0=10^{-3}$ (black) and $\epsilon_0=1$ (blue). 
    The dust layer thickness is fixed to $\Hd=0.99\Hg$ so the 
    dust-to-gas ratio $\epsilon$ is approximately constant with
    height. Vertical buoyancy here is due to dust-loading. 
    \label{compare_vshear_fixHd}
    }
\end{figure}

@@

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably, surface modes, which are typically fastest growing in
the dust-free case, are suppressed in dusty disks for $\epsilon_0\geq
0.1$ (i.e. $\epsilon_0> \hgas$).   
The body modes' growth rates remain $ O(h_\mathrm{g}\OmK)$ 
but their oscillation frequency increases with
dust-loading, i.e. it increases with the vertical buoyancy. 
The total number of modes do not change 
significantly. This is in contrast with the effect of increasing
cooling times in an adiabatic gas disk, for which \citetalias{lin15}
find fewer unstable modes. 
%until the system is eventually completely stable.

@@

We analyze two disks with $\Hd=0.1\Hg$ and 
$\Hd=0.99\Hg$. Fig. \ref{compare_vshear_fixZ} compares the vertical
shear  rate and buoyancy frequency. For $|z|\gtrsim 0.4\Hg$ the two
disks have the same profile with vertical shear dominating over
buoyancy. We thus expect perturbations away from the disk midplane in 
both cases. For $|z|\lesssim 0.4\Hg$, a thin dust 
layer with $\Hd=0.1\Hg$ boosts the vertical shear rate, but the
associated buoyancy is larger still, implying the mid-plane should be
stable. 
%Thus we expect the mid-plane of 
%the disk with $\Hd=0.1\Hg$ to have limited vertical motions.

@@

Fig. \ref{result2d_fixZ} compares the fastest growing VSI body modes  
with $k_x\Hg=30$ for the two cases above.\citepalias[The thinner domain
  adopted here eliminates surface modes, ][]{lin15}.   
We find very similar mode 
frequencies 
\begin{align*}
  \sigma = \begin{cases}
    \left(0.3053\ii - 0.8142\right)h_\mathrm{g}\OmK & \Hd=0.99\Hg, \\
    \left(0.3178\ii - 1.2237\right)h_\mathrm{g}\OmK & \Hd=0.1\Hg,
  \end{cases}
\end{align*}
since the vertical shear profile is similar throughout most of the
disk. However, meridional motions are suppressed near the midplane of
the $\Hd=0.1\Hg$ disk, as expected from the larger buoyancy frequency
relative to vertical shear there. This leads to a structure
analogous to PPD dead zones: a quiescent midplane between active
surface layers \citep{gammie96}.

@@

To connect with similar studies, here  we also use 
the Richardson number $\rich \equiv N_z^2/\left(r\p_z\Omega\right)^2$
to label calculations \citep{youdin02}.   
Numerical simulations show    
non-axisymmetric instabilities develops when $\rich\lesssim0.1$, brought about
by very thin dust layers \citep{chiang08, lee10}. We show that \emph{axisymmetric}
instabilities never develop in radially uniform disks, however small $\rich$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_thindust} 
  \caption{Maximum VSI growth rate for ultra-thin dust layers $\Hd 
    \leq 0.01\Hg$ (with radially uniform dust-to-gas  ratio). Vertical shear is dominated by that due to 
    vertical dust stratification, but axisymmetric instability is still 
    associated with the radial temperature gradient $q$. 
 Here, $\rich_0$ is the minimum 
    Richardson number in the domain in the limit $q\to0$. The disk is stable to axisymmetric perturbations 
    in the strictly isothermal limit, regardless of the dust layer thickness. 
    \label{ultra_thin}
    }
\end{figure}

@@

In \S\ref{iso_perfect} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.  
To demonstrate this numerically we consider the
same disk as above but with $q=0$ and choose $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. Then vertical shear $\p_z\Omega$ is 
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).

We show unstable modes for this disk in
Fig. \ref{vert_mixed_modes}. In order to obtain appreciable growth   
rates, we consider super-Solar metalicity $Z=0.03$ and high
radial wavenumber $k_x\Hg=1800$. As expected from the discussion in
\S\ref{iso_perfect}, the disk admits purely growing modes with
$\omega=0$. This is distinct from the growing oscillations associated
with classic VSI discussed above.

### 2017-08-08 summary.tex
misc home edits
final revision before submission, spell-checked

%\subsubsection{Locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
polytropic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

@@

\subsubsection{Classic gravitational instabilities} %vertical structure 
The addition of dust enhances gravitational
instability (GI) because dust particles contribute to the
total disk mass but not thermal pressure, which effectively lowers the
disk temperature \citep[][]{thompson88,shi13}. For typical dust-loading 
$\epsilon, Z\ll1$, this effect is unimportant. However, if $\epsilon$ is
large (e.g. due to dust settling) then the reduced temperature
$\widetilde{T}$ may be lowered to enable instability.

@@

Another potential connection to previous results for gas disks is  
gravito-turbulence. Cooling, self-gravitating gaseous disks 
sustain a turbulent state where shock heating due to gravitational 
instabilities are balanced by radiative cooling 
\citep{gammie01}. Since dust-gas drift appears as a diffusion or 
cooling in our framework (Eq. \ref{eff_energy}), it may be conceivable
to have `dusty gravito-turbulence'. As self-gravity increases the
local density, the associated pressure maxima attract dust-particles,
but the back-reaction onto the gas may try to flatten the pressure
bump \citep{taki16}.

@@

\subsubsection{Rossby wave instability}
The Rossby wave instability (RWI) is a non-axisymmetric, 2D shear
instability that operates in thin disks when it has radial structure
\citep{lovelace99,li00}. These studies consider adiabatic pure gas and 
show instability is possible if there is an extremum in the generalized
potential vorticity 
%\begin{align}
 $\mathcal{V}_\mathrm{g} =
 \kappa^2\mathcal{S}_\mathrm{g}^{-2/\gamma}/2\Omega\Sigma_\mathrm{g}$, 
%\end{align} 
where $\mathcal{S}_\mathrm{g} = P/\Sigma_\mathrm{g}^\gamma$. Here $P$
should be interpreted as the vertically integrated pressure. The
non-linear result of the RWI is vortex formation \citep{li01}.

@@

Since the necessary physical ingredient for the ZVI is vertical
buoyancy, it requires an almost purely adiabatic gas ($\tcool\OmK\gg
1$). This limits its applicability in typical protoplanetary disks to 
$\lesssim 1$AU \citep{lesur16, malygin17}. However, this consideration
assumes a dust-free disk as far as the dynamics is concerned.

On the other hand, we have show that dust-loading induces an effective
buoyancy even in isothermal disks. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce `dusty ZVI'. This may allow the zombie
vortices to develop $\gtrsim 1$AU in PPDs.

@@

\section{Summary}\label{summary}
In this paper we develop the analogy between isothermal 
dusty-gas and adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy being conserved
following a adiabatic, pure gas.

@@

We explicitly show that for a fixed gas equation of state, the  
evolutionary equation for the dust-to-gas ratio $\rhod/\rhog$ may be replaced by an 
effective energy equation. This leads to a 
natural definition of the effective entropy of isothermal dusty gas as  
\begin{align*}
  \seff  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
The effect of finite dust-gas friction appears as an energy
source term.  This analogy with standard 
hydrodynamics with cooling/heating allow us to find dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

@@

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show in SI the evolution of gas
pressure lags behind dust density. This is in fact a general property of 
instabilities driven by dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the dusty-gas mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work that amplifies 
oscillations. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

@@

>>>>

%\subsubsection{Locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
polytropic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right). 
%+ \mathcal{C},\\
%  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
%  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

@@

\subsubsection{Classic gravitational instabilities} %vertical structure 
The addition of dust enhances gravitational
instability (GI) because dust particles contribute to the
total disk mass but not thermal pressure, which effectively lowers the
disk temperature \citep[][]{thompson88,shi13}. For typical dust-loading 
$\tepsilon, Z\ll1$, this effect is unimportant. However, if $\epsilon$ is
large (e.g. due to dust settling) then the reduced temperature
$\widetilde{T}~=~T(1 - \tepsilon)$ may be lowered to enable instability.

@@

Another potential connection to previous results for gas disks is  
gravito-turbulence. Cooling, self-gravitating gaseous disks 
sustain a turbulent state where shock heating due to gravitational 
instabilities are balanced by radiative cooling 
\citep{gammie01}. Since dust-gas drift appears as a diffusion or 
cooling in our framework (Eq. \ref{eff_energy}), it may be conceivable
to have `dusty gravito-turbulence'. As self-gravity increases the
local density, the associated pressure maxima attract dust-particles,
but the back-reaction onto the gas may try to flatten the pressure
bump \citep{taki16}, thus enabling a quasi-steady state.

@@

\subsubsection{Rossby wave instability}
The Rossby wave instability (RWI) is a non-axisymmetric, 2D shear
instability that operates in thin disks when it has radial structure
\citep{lovelace99,li00}. These studies consider adiabatic pure gas and 
show instability is possible if there is an extremum in the generalized
potential vorticity 
%\begin{align}
 $\mathcal{V}_\mathrm{g} =
 \kappa^2\mathcal{S}_\mathrm{g}^{-2/\gamma}/2\Omega\Sigma_\mathrm{g}$, 
%\end{align} 
where $\mathcal{S}_\mathrm{g} = P/\Sigma_\mathrm{g}^\gamma$ is essentially
the gas entropy. Here $P$
should be interpreted as the vertically integrated pressure. The
non-linear result of the RWI is vortex formation \citep{li01}.

@@

Since the necessary physical ingredient for the ZVI is vertical
buoyancy, it requires an almost purely adiabatic gas with long cooling times ($\tcool\OmK\gg
1$). This limits its applicability in typical protoplanetary disks to 
$\lesssim 1$AU \citep{lesur16, malygin17}. However, this consideration
assumes a dust-free disk as far as the dynamics is concerned.

On the other hand, we have show that dust-loading induces an effective
buoyancy even in isothermal disks. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading (specifically a vertical gradient in the dust-to-gas ratio), 
and thus produce `dusty ZVI'. This may allow the zombie 
vortices to develop $\gtrsim 1$AU in PPDs.

@@

\section{Summary}\label{summary}
In this paper we develop the analogy between isothermal 
dusty-gas and pure ideal gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, with stopping time $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy conservation 
following an adiabatic, pure gas.

@@

We explicitly show that for a fixed gas equation of state, the  
evolutionary equation for $\rhod/\rhog$ may be replaced by an 
effective energy equation. This leads to a 
natural definition of the effective entropy of isothermal dusty gas as  
\begin{align*}
  \seff  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)},  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
which implies that a non-uniform dust-to-gas ratio induces buoyancy forces.  
The effect of finite dust-gas friction appears as an energy
source term.  This analogy with standard  
hydrodynamics with cooling/heating allow us to find dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

@@

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the gaseous vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We 
explicitly show in SI the evolution of gas 
pressure lags behind dust density. In fact, this is a general property of 
overstabilities driven by dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the dusty-gas mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work that amplifies 
oscillations. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

@@

The dusty/adiabatic gas equivalence also offers a simple way to simulate dusty protoplanetary disks 
using purely hydrodynamic codes. All that is required is a re-interpretation of the fluid variables and additional 
source terms in the usual energy equation. The latter is already available in many public codes. In a follow-up work
we will apply this approach to study dusty disk-planet interaction.
